Peaks Over Threshold for Bursty Time Series

### 2018-04-16 article_springer/article_springer.Rmd
defer derivation of scaling limit to appendix

---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  message = FALSE,
  fig.height = 3, out.width = '\\textwidth'
  )
library(CTRM)
```

@@

We review the CTRM, or max-renewal model, in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Sections 3 &
4. 
A statistical procedure to estimate model parameters via stability plots 
is outlined in Section 5, and diagnostic plots for model criticism are 
discussed in Section 6. 
In Section 7, we give case studies for simulated data as well as for a
real-world dataset, with a discussion in Section 8. 
All statistical computations have been done in R, and all code, data, as 
well as the source code for this manuscript has been organized into an R 
software package, available at <https://github.com/strakaps/CTRM>.

# CTRMs

![\label{fig:ctrm} A CTRM sample path.](just-tikz-figure0.pdf)

Continuous Time Random Maxima (CTRM), or Max-Renewal processes, track the 
largest event in a renewal sequence of events [@Basrak2014; 
@MeerschaertStoev08; @Hees16]:

**Definition (CTRM):**

: Let $(W,J), (W_1, J_1), (W_2, J_2), \ldots$ be i.i.d. pairs of random 
variables, where the $W_k > 0$ are interpreted as the *waiting times* or 
inter-arrival times and $J_k \in [x_L, x_R]$ as the *event magnitudes* 
or marks
($x_L \in [-\infty, +\infty), x_R \in (-\infty, +\infty]$). 
The renewal process associated with the $W_k$ is 
$$N(t) = \max\{n: W_1 + \ldots + W_n \le t\},$$
and the running maximum is $M(n) := \bigvee_{k=1}^n J_k$. Then the process
$$V(t)=M(N(t)) = \bigvee_{k=1}^{N(t)} J_k, \quad t \ge 0.$$
is called a CTRM (Continuous Time Random Maxima) process. 
If $W$ and $J$ are independent, the CTRM is said to be *uncoupled*. 
We let $M(0) = x_L$.

Figure \ref{fig:ctrm} shows a CTRM sample path.
The alternative case where one considers the maximum of the first 
$(N(t)+1)$'s event is called an *OCTRM process* (Overshooting Continuous 
Time Random Maxima; or *Model II* by @Sumita1983). 
The conceptual difference between CTRM and OCTRM is the order of waiting 
times and events, which is $W_1,J_1,W_2,J_2,..$ for the CTRM and 
$J_1,W_1,J_2,W_2,...$ for the OCTRM. 
In the uncoupled case, unsurprisingly CTRM and OCTRM are asymptotically 
equivalent. 
In the coupled case, however, their distributions 
can be dramatically different [@Hees16].

The main purpose of this article is to introduce and discuss a method for 
fitting data to an uncoupled CTRM, based on the following observable data:

@@

: The random variables
$$X(\ell) = V(T(\ell)) - \ell, \quad 
T(\ell) = \inf\{t: V(t) > \ell\}, 
\quad \ell \in (x_L, x_R)$$
are called the *exceedance* resp. *time until exceedance* of level $\ell$.

#### POT-like approach.
Since we assume the uncoupled case, $X(\ell)$ and $T(\ell)$ are 
independent. 
To see this, note that $X(\ell)$ is, in distribution, simply equal to
$J | J > \ell$, independent of any waiting time $W_k$. 
For our inference on the CTRM model, we extract realizations of 
$X(\ell)$ and $T(\ell)$ from time series data by considering thresholds 
$\ell$ ranging over some interval $[\ell_0, x_R)$. 
(A reasonable default value for $\ell_0$ is the $95$-th
percentile of $J$.)
Write $X(\ell,i)$ for the $i$-th exceedance of level $\ell$, 
and $T(\ell,i)$ for the inter-arrival time between
$X(\ell,i-1)$ and $X(\ell,i)$. 
Thus for any threshold $\ell \in [\ell_0, x_R)$,
we construct sequences $\{T(\ell,i)\}_{i}$ and 
$\{X(\ell,i)\}_{i}$. 
These sequences are i.i.d. because a renewal occurs at each threshold
crossing time, and independent because $W_k$ and $J_k$ are uncoupled.
Extracting the exceedances $X(\ell)$ for inference is a standard procedure
known as the Peaks Over Threshold (POT) approach in extreme value theory
[@beirlantBook].
Figure \ref{fig:thresholdedBursty} illustrates the exceedances 
$X(\ell, i)$ in red and the times until exceedance $T(\ell, i)$ as the 
preceding intervals marked by blue crosses.

```{r thresholdedBursty, message=FALSE, fig.cap="\\label{fig:thresholdedBursty}Exceedances (red) and Times until Exceedance (durations between blue crosses) for a given threshold $\\ell$ (dashed line)."}
#rpareto <- function(n,tail){
#          if(tail<=0) stop("tail must be positive")
#          rp <- (runif(n)/gamma(1-tail))^(-1/tail)
#          rp}
tail <- params$tail
n <- params$n 
sigma <- (cos(pi*tail/2))^(1/tail)
times <- cumsum(stabledist::rstable(n, tail, 1, sigma, pm=1))
#times<-cumsum(rpareto(n,tail))
magnitudes <- extRemes::revd(n, scale = 1, shape = 0)
sim_ctrm <- ctrm(data.frame(times, magnitudes))
plot(sim_ctrm, p = 0.005)
```

@@

>>>>

preamble: |
  \usepackage{amssymb}
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  message = FALSE,
  fig.height = 3, out.width = '\\textwidth'
  )
library(CTRM)
library(magrittr)
```

@@

We review the marked renewal process in Section 2, and 
derive a scaling limit theorem for inter-exceedance times in Sections 3 &
4. 
A statistical procedure to estimate model parameters via stability plots 
is outlined in Section 5, and diagnostic plots for model criticism are 
discussed in Section 6. 
In Section 7, we give case studies for simulated data as well as for a
real-world dataset, with a discussion in Section 8. 
All statistical computations have been done in R, and all code, data, as 
well as the source code for this manuscript has been organized into an R 
software package, available at <https://github.com/strakaps/CTRM>.

# Marked Renewal Process (MRP)

@@

: Let $(W,J), (W_1, J_1), (W_2, J_2), \ldots$ be i.i.d. pairs of random 
variables, where the $W_k > 0$ are interpreted as the *waiting times* or 
inter-arrival times and $J_k \in [x_L, x_R]$ as the *event magnitudes* 
or marks
($x_L \in [-\infty, +\infty), x_R \in (-\infty, +\infty]$). 
If $W$ and $J$ are independent, the MRP is said to be *uncoupled*.

In this setup, the $k$ magnitude $J_k$ occurs at time
$T_k = W_1 + \ldots + W_k$. 
Based on this simple model, we derive the following two random variables:

**Definition (MRP):**

: Given a threshold $\ell \in (x_L, x_R)$, 
consider the stopping time 
$$\tau(\ell) := \min\{k: J_k > \ell\},\quad \ell \in (x_L, x_R).$$
The random variables
$$X(\ell) = J_{\tau(\ell)} - \ell, \quad 
T(\ell) = \sum_{k=1}^{\tau(\ell)} W_k$$
are called the *exceedance* resp. *time until exceedance* of level $\ell$.

By restarting the MRP at $\tau(\ell)$, one can inductively construct the
two i.i.d. sequences $T(\ell,n)$ and $X(\ell, n)$, $n \in \mathbb N$.
Moreover, since we assume the uncoupled case, the two sequences are independent.[^1]

[^1]: (To see why, note that $X(\ell)$ is, in distribution, simply equal to 
  $J | J > \ell$, independent of any waiting time $W_k$.)

```{r thresholdedBursty, message=FALSE, fig.height=7, fig.cap="\\label{fig:thresholdedBursty}Exceedances (red) and Times until Exceedance (durations between blue crosses) for a given threshold $\\ell$ (dashed line)."}
#rpareto <- function(n,tail){
#          if(tail<=0) stop("tail must be positive")
#          rp <- (runif(n)/gamma(1-tail))^(-1/tail)
#          rp}
tail <- params$tail
n <- params$n 
sigma <- (cos(pi*tail/2))^(1/tail)
times <- cumsum(stabledist::rstable(n, tail, 1, sigma, pm=1))
#times<-cumsum(rpareto(n,tail))
magnitudes <- extRemes::revd(n, scale = 1, shape = 0)
sim_ctrm <- ctrm(data.frame(times, magnitudes)) 
par(mfrow = c(2,1))
plot(sim_ctrm, p = 0.01, main = "Simulated MRP") 
flares %>% ctrm() %>% plot(p = 0.02, log = 'y', main = "HXRBS data") 
```

Figure \ref{fig:thresholdedBursty} shows a simulated dataset in the top 
panel, where $W$ has a stable distribution with tail parameter 
$\beta =$ `r params$tail` (and skewness $1$ and location $0$), and 
where $J$ is from a standard Gumbel distribution. In the bottom panel, 
we plot a time series of solar flare intensities derived from a NASA
dataset [@HXRBS][^*]. Clearly, the simulated data exhibit long intervals 
_without any_ events, whereas in the real-world dataset events appear 
continuously. The threshold exceedances, however, appear to have the same 
statistical behaviour in both models. Observations below a threshold are 
commonly discarded in Extreme Value Theory; the MRP models 
interprets these observations as noise and filters them out.

[^*]: The "complete Hard X Ray Burst Spectrometer event list" 
is a comprehensive reference for all measurements of the Hard X 
Ray Burst Spectrometer on NASA's Solar Maximum Mission from the time of 
launch on Feb 14, 1980 to the end of the mission in Dec 1989. 12,776 
events were detected, with the "vast 
majority being solar flares". The list includes the start time, peak time, 
duration, and peak rate of each event. We have used "start time" as the 
variable for event times, and "peak rate" as the variable for event 
magnitudes.

@@

\appendix

For our inference on the CTRM model, we extract realizations of 
$X(\ell)$ and $T(\ell)$ from time series data by considering thresholds 
$\ell$ ranging over some interval $[\ell_0, x_R)$. 
(A reasonable default value for $\ell_0$ is the $95$-th
percentile of $J$.)
Write $X(\ell,i)$ for the $i$-th exceedance of level $\ell$, 
and $T(\ell,i)$ for the inter-arrival time between
$X(\ell,i-1)$ and $X(\ell,i)$. 
Thus for any threshold $\ell \in [\ell_0, x_R)$,
we construct sequences $\{T(\ell,i)\}_{i}$ and 
$\{X(\ell,i)\}_{i}$. 
These sequences are i.i.d. because a renewal occurs at each threshold
crossing time, and independent because $W_k$ and $J_k$ are uncoupled.
Extracting the exceedances $X(\ell)$ for inference is a standard procedure
known as the Peaks Over Threshold (POT) approach in extreme value theory
[@beirlantBook].
Figure \ref{fig:thresholdedBursty} illustrates the exceedances 
$X(\ell, i)$ in red and the times until exceedance $T(\ell, i)$ as the 
preceding intervals marked by blue crosses.
